---
title: "HW2"
author: "RJ Cass"
format: pdf
---

```{r init}
suppressPackageStartupMessages(library(dplyr))
dt <- read.table('cult_data.dat', header = FALSE)
colnames(dt) <- c('field', 'cultivar', 'soil', 'weight')
dt <- dt %>% 
    group_by(soil, cultivar) %>% 
    mutate(group = cur_group_id()) %>% 
    ungroup()
dt <- dt %>% 
    mutate(across(c(field, cultivar, soil, group), as.factor))
```

# 1

```{r}
anova(lm(weight~cultivar*soil, data = dt))
```

According to the ANOVA output, when testing just the cultivar and soil effects (and their interaction), cultivar and the interaction do not have a significant effect on the growth of the plants. However, we do see that the soil has a significant effect. 

# 2

```{r}
y <- dt$weight
w <- model.matrix(y~group-1,data=dt)
muhat <- solve(t(w)%*%w)%*%t(w)%*%y
muhat
# Group 1: Control, A
# Group 2: Control, B
# Group 3: Dead, A
# Group 4: Dead, B
# Group 5: Live, A
# Group 6: Live, B
```


# 3

```{r full_contrast}

y <- dt$weight
w <- model.matrix(y ~ group-1, data=dt)
x <- model.matrix(y ~ group, data=dt)

#              | Cultivar a  | Cultivar b |
# Soil Control |   1         |     2      |
# Soil Dead    |   3         |     4      |
# Soil Live    |   5         |     6      |


c <- rbind(
  c(1/6,1/6,1/6,1/6,1/6,1/6), # Overall mean
  c(1/2,1/2,-1/2,-1/2,0,0), # Soil Control vs. Soil Dead
  c(0,0,1/2,1/2,-1/2,-1/2), # Soil Dead vs. Soil Live
  c(1/3,-1/3,1/3,-1/3,1/3,-1/3), # Cultivar A vs. B
  c(1,-1,-1,1,0,0), # Interaction 1
  c(0,0,1,-1,-1,1) # Interaction 2
)

muhat <- solve(t(w)%*%w)%*%t(w)%*%y
beta <- solve(t(x)%*%x)%*%t(x)%*%y
sse <- t(y-x%*%beta)%*%(y-x%*%beta)
mse <- as.numeric(sse/(24-6))
varmuhat <- mse*solve(t(w)%*%w)
```

```{r}
c_cult <- c[4, ]
fhat <- t(c_cult%*%muhat)%*%solve(c_cult%*%varmuhat%*%c_cult)%*%c_cult%*%muhat/1
fhat
1-pf(fhat,1,23)
```

No, this test is not significant

# 4 

```{r}
c_soil <- c[2:3, ]
fhat <- t(c_soil%*%muhat)%*%solve(c_soil%*%varmuhat%*%t(c_soil))%*%c_soil%*%muhat/2
fhat
1-pf(fhat,2,22)
```

Yes, this test is significant

# 5 

```{r}
c_int <- c[5:6, ]
fhat <- t(c_int%*%muhat)%*%solve(c_int%*%varmuhat%*%t(c_int))%*%c_int%*%muhat/2
fhat
1-pf(fhat,1,22)
```

No, there are not significant interactions

# 6

```{r}
dt_red <- dt[!(dt$cultivar == 'b' & dt$soil == 'liv'),]

w <- model.matrix(weight ~ group-1, data=dt_red)
w1 <- w[,1:5] # Remove the all 0s column
ym <- dt_red$weight
muhat <- solve(t(w1)%*%w1)%*%t(w1)%*%ym
muhat
```

# 7

```{r}
ai <- solve(c)
Xm <- w%*%ai
# Remove the second interaction (the one that depends on group 6) row to make it full rank
Xn <- Xm[,-6]

betahat <- solve(t(Xn)%*%Xn)%*%t(Xn)%*%ym
sse <- t(ym-Xn%*%betahat)%*%(ym-Xn%*%betahat)
mse <- as.numeric(sse)/(20-5)
varbeta <- mse*solve(t(Xn)%*%Xn)
newmuhat <- ai[,-6]%*%betahat
newmuhat
```

The estimate for the cell is 34.55, while the original with data was 33.525, so the estimate was pretty close (only ~ 1 away).

# 8

```{r}
vestmu <- ai[,-6]%*%varbeta%*%t(ai[,-6])
cbind(newmuhat,sqrt(diag(vestmu)))
```

Standard error of the estimated cell mean is ~1.13, while the rest of the errors are about half that much, which makes sense. The error of the cell we estimated should be larger than the other estimates. 